### 2015-04-20 rapport/formalism.tex
rapport_200415

A mesh $\mathcal{M}$ defines the discretization of the continuous space domain $\Omega$ of a set of PDEs.

\medskip
\begin{mydef}
\textit{A mesh is a connected undirected graph without bridges.}
\end{mydef}

$E_i$ denotes a set of elements constructed by a function $elem_i$ which defines a precise association of nodes and edges of a mesh. If we denote $nodes$ the function to get nodes of a mesh and $edges$ the same for edges, the funtion $elem_i$ is defined as
\begin{equation*}
elem_i : nodes(\mathcal{M}) \times edges(\mathcal{M}) \rightarrow E_i
\end{equation*}
For example, the set of cells $E_0$ in a Cartesian 2D mesh could be defined by exactly four nodes and four edges connected as a cycle. But we could also define another set of elements $E_1$ as the simple set of nodes etc.

A mesh can be structured (as Cartesian or curvilinear meshes), unstructured, regular or irregular (without the same topology for each element) and hybrid if mixing those.

@@

The discretization of the continuous time domain $\mathcal{T}$ is denoted $T$. $T$ is responsible for the iteration time steps of the numerical simulation.

In a numerical simulation a set of data, or quantities, are applied onto the mesh and represent the set of values to compute, or to use, for computation. We denote the set of data applied on the mesh by $\Delta$, such that $\delta \in \Delta$ is a function which associates each element $el \in E_i$ to a value $v \in V$, 
\begin{equation*}
\delta : E_i \rightarrow V
\end{equation*}
One can notice that in applied mathematics, the signature of $\delta$ would be $\delta : E_i \times T \rightarrow V$, however when programming a numerical simulation it is not wise to store all values of each time iteration.

Finally, in a numerical simulation are performed a set of ordered numerical computations denoted $\Gamma$. A numerical computation in $\Gamma$ is denoted 
\begin{equation*}
c(R,w,D,e). 
\end{equation*}
In this definition $R \subset \Delta, w \in \Delta$ are respectively the data read and written to compute the numerical expression $e$ (a single data is written in a single computation). $D$ is one of the subsets $E_i \subset \mathcal{M}$. It has to be noticed that at each time iteration, all the elements of a mesh are computed. However, it happens that the computation of the mesh elements is splitted in different computations (for example the computation of the physical border). In this case additional $E_i$ can be specified for the mesh $\mathcal{M}$.
The numerical expression $e$ computes the data $w$ from a set of data $R$.
\begin{equation*}
e : R \rightarrow w
\end{equation*}

A \textit{multi-stencil program} is defined by the quadruplet
\begin{equation}
\mathcal{MSP}(T,\mathcal{M},\Delta,\Gamma),
\end{equation}

If the number of computations in $\Gamma$ is $card(\Gamma)=m$, such that $\bigcup_{i=0}^{m-1}c_i = \Gamma$, then $\bigcup_{i=0}^{m-1}R_i \cup w_i \subseteq \Delta$.

A computation $c \in \Gamma$ can be of two different types. The first type is called a \textit{stencil computation} and involves an additional information denoted $\mathcal{N}$ which represents a neighborhood function around an element
\begin{equation}
\mathcal{N} : E_i \rightarrow E_k \times E_k \times ...
\end{equation}
, where $E_k$ is a set of elements of $\mathcal{M}$ which could be equal to $E_i$ or not. In other words, from an element $el \in E_i$ the function $\mathcal{N}$ returns a set of elements of $E_k$ which represents the neighborhood of $el$. The function $\mathcal{N}$ is also sometimes called the \textit{stencil shape}, or the \textit{stencil} in applied mathematics.

A \textit{stencil computation} is a quintuplet
\begin{equation}
s(R,w,D,e,\mathcal{N}),
\end{equation}
where $R \subset \Delta, w \in \Delta$ are read and written in the numerical expression $e$, and $D$ is one the subsets $E_i$. In a stencil computation $s$, $\forall d \in D$, the stencil numerical expression $e$ is applied such that $w(d) = e(R(d),R(\mathcal{N}(d))$. In this work, a stencil computation $s(R,w,D,e,\mathcal{N})$ always verifies $R \cap w = \emptyset$, otherwize an implicit numerical scheme has to be solve which is over the scope of this paper.

Figure~\ref{} gives an example of a stencil computation $s(R,w,D,e,\mathcal{N})$, on a two dimensional Cartesian mesh, where $R=\{A,B\}$, $w=C$, and where for $(x,y) \in D$
\begin{equation*}
e(R(x,y),R(\mathcal{N}(x,y)) = A(x,y)+B(x,y+1)+B(x,y-1)+B(x+1,y)+B(x-1,y).
\end{equation*}

Finally, the second type of numerical computation is a local computation where $e$ does not involve a neighborhood function $\mathcal{N}$
\begin{equation}
l(R,w,D,e)
\end{equation}

%-------------------------------------
\subsection{Parallelization}
\label{sect:parall}
% Multi-stencil mesh-based numerical simulation can be parallelized in various ways and is an interesting kind of application to take advantage of modern heterogeneous HPC architectures, mixing clusters, multi-cores CPUs, vectorization units, GPGPU and many-core accelerators.

% Maybe the more important parallelization technique of such simulations is called \textit{SPMD} (Single Program, Multiple Data) which can also be called \textit{coarse-grain data parallelism}. This concept represents the well-known and broadly-used domain decomposition technique to split the mesh in balanced sub-meshes for available resources. Because of the good locality degree of numerical simulations, only a small number of additional elements, called a \textit{recorvery} or a \textit{ghost} area, are needed to exchange information computed by another resource to correctly compute stencils. As the amount of communications of such parallelization are rather small compared to the amount of computations (small neighborhood), and also by the addition of recovery of communications by computations, this parallelization technique has proved many times its efficiency and scalability on numerical simulations~\cite{}. This parallelization technique can be applied on shared and distributed memory architectures (clusters, multi-core CPUs), and it stays the most efficient parallelization technique if the numerical simulation has to be executed on a distributed memory architecture. For this reason, this parallelization stays a reference in the domain and is still studied for more complex simulations, using unstructured meshes, multi-grids techniques etc.

% However, because of energy consumption constraints, most of the \textit{top500}~\footnote{http://www.top500.org} supercomputer centers are not only composed of clusters of multi-core machines, but also of accelerator cards such as GPGPU and many-cores. In addition to this, vectorization units are also available on most CPUs. This introduces another kind of parallelization, linked to the first one, called in this paper \textit{fine-grain data parallelism}, which is composed of \textit{SIMD} (Single Instruction, Multiple Data) and \textit{SIMT} (Single Instruction, Multiple Threads) parallelization techniques. Such techniques typically compute a numerical expression in parallel on a vector, using vectorization units or threads, which perfectly matches the type of computation in the nested loops of numerical simulations.

% Finally, the last parallelization technique which can be applied on various architectures (clusters, multi-cores, many-cores, GPGPU), is the \textit{task parallelism} technique. Instead of spliting the data of the simulation (or the vector used in a numerical expression), the different tasks of the simulation are identified and scheduled over the available resources during the execution (at runtime)~\cite{}. As numerical computations of a simulation can be considered as tasks, linked together through data dependencies for example, this parallelization technique can be used for HPC numerical simulations too.

% This section discussed the various parallelization techniques which can be used for numerical simulations. As a result, it seems that, by a combination of existing techniques, modern heterogeneous parallel architectures can be fully harnessed by such applications~\cite{}. In this paper is presented a component-based model to expose those three kinds of parallelism in the overall simulation, while hiding them from the user (numericians).

% %-------------------------------------
% \subsection{Dependencies and synchronizations}
% \label{sect:dep}
% To define and explain the component-based implicit parallelism model presented in this paper, two additional definitions handle the relations between data read and written into computations. Those relations are responsible for dependencies and synchronizations.

% \medskip
% \begin{mydef}
% \textit{For two computations $c_1(R_1,w_1,D_1,e_1)$ and $c_2(R_2,w_2,D_2,e_2)$, it is said that $c_2$ is dependant from $c_1$, denoted $c_1<c_2$, if $w_1 \cap R_2 \neq \emptyset$. In this case, $c_1$ has to be computed before $c_2$. The binary relation $c_1<c_2$ represents a \textit{dependency}.}
% \end{mydef}

% \medskip
% Dependencies lead to one computation before another, but when the program is parallelized some dependencies are also responsible for synchronizations between different processes. The opposite definition can also be given,

% \medskip
% \begin{mydef}
% \textit{For two computations $c_1(R_1,w_1,D_1,e_1)$ and $c_2(R_2,w_2,D_2,e_2)$, it is said that $c_2$ is independant from $c_1$, denoted $c_1 \not <c_2$ or $c_1 \parallel c_2$, if $w_1 \cap R_2 = \emptyset$.}
% \end{mydef}

% Considering a coarse-grain data parallelized multi-stencil program $\mathcal{MSP}$,

% \medskip
% \begin{mydef}
% \textit{A depedency between two computations $c_1<c_2$ is a \textit{synchronisation}, denoted $c_1 \ll c_2$, if $c_2=s_2(R_2,w_2,D_2,e_2)$, and $w_1 \cap R_2 \neq \emptyset$.}
% \end{mydef}

% \medskip
% Actually, in a parallel stencil program, synchronizations between processes are always needed for the set of data $R$ of a stencil $s(R,w,D,e)$ because the expression $e$ involves a neighborhood $\mathcal{N}$ which could be handled by another processor or resource.

>>>>

A mesh $\mathcal{M}$ defines the discretization of the continuous space domain $\Omega$ of a set of PDEs and is defined as followed.

\begin{mydef}
\textit{A mesh is a connected undirected graph $\mathcal{M}=(V,E)$, where $V$ is the set of vertices and $E$ the set of edges. The set of edges $E$ of a mesh $\mathcal{M}=(V,E)$ does not contain bridges.}
\end{mydef}

\begin{mydef}
$D_i$ is a set of elements of a mesh $\mathcal{M}=(V,E)$, constructed by a function $domain_i$ which defines a precise association between $V$ and $E$, $domain_i : V \times E \rightarrow D_i$.
\end{mydef}
For example, the set of cells $D_0$ in a Cartesian 2D mesh could be defined by exactly four vertices and four edges connected as a cycle. But we could also define another set of elements $D_1$ as the simple set of vertices $V$ etc.

A mesh can be structured (as Cartesian or curvilinear meshes), unstructured, regular or irregular (without the same topology for each element) and hybrid.

@@

\begin{mydef}
The discretization of the continuous time domain $\mathcal{T}$ is denoted $T$ such that $\forall t_i, t_{i+1} \in T, \exists \Delta t \in \mathbb{R}$, $t_{i+1} = t_i + \Delta t$. Thus, $T$ is responsible for the iteration time steps of the numerical simulation. 
\end{mydef}

In a numerical simulation a set of data, or quantities, are applied onto the mesh and represent the set of values to compute, or to use, for computation.

\begin{mydef}
The set of data applied on the mesh is denoted by $\Delta$, such that $\delta \in \Delta$ is a function which associates each element $d \in D_i$ to a value $v \in V$, $\delta : D_i \rightarrow V$.
\end{mydef}
One can notice that in applied mathematics, the signature of $\delta$ would be $\delta : E_i \times T \rightarrow V$, however when programming a numerical simulation it is not wise to store all values of each time iteration.

\begin{mydef}
A numerical expression $\text{exp}$ is a function which represents how to compute, for an element $d \in D_i$, a data $w \in \Delta$ (written data) with a set $R \subset \Delta$ of input data (read data), $\text{exp} : R \times D_i \rightarrow w \times D_i$.
\end{mydef}

\begin{mydef}
A computation $c$ of a numerical simulation is defined as $c(R,w,D,\text{exp})$, where $R \subset \Delta, w \in \Delta$ and $\text{exp}$ a numerical expression $\text{exp} : R \rightarrow w$. $D$ is one of the subsets $D_i \subset \mathcal{M}$, such that $w : D \rightarrow V$.
\end{mydef}
It has to be noticed that at each time iteration, all the elements of a mesh are computed. However, it happens that the computation of the mesh elements is splitted in different computations (for example the computation of the physical border). In this case additional $D_i$ can be specified for the mesh $\mathcal{M}$.

\begin{mydef}
The set of $n$ ordered computations of a numerical simulation is denoted $\Gamma = [c_i]_{0 \leq i \leq n-1}$, such that $\forall c_i,c_j$ with $i \leq j$, $c_i$ is computed before $c_j$, and $c_j$ can be computed only when $c_i$ is finished.
\end{mydef}

\begin{mydef}
Finally, a \textit{multi-stencil program} is defined by the quadruplet $\mathcal{MSP}(T,\mathcal{M},\Delta,\Gamma)$.
\end{mydef}
%If the number of computations in $\Gamma$ is $card(\Gamma)=n$, such that $\bigcup_{i=0}^{n-1}c_i = \Gamma$, then $\bigcup_{i=0}^{m-1}R_i \cup w_i \subseteq \Delta$.

\begin{mydef}
The neighborhood $\mathcal{N}$ of an element $d \in D_i$ is a function to define a set of elements in any $D_k \subset \mathcal{M}$, $\mathcal{N} : D_i \rightarrow D_k \times D_k \times \dots$.
\end{mydef}
The function $\mathcal{N}$ is also sometimes called the \textit{stencil shape}, or the \textit{stencil} in applied mathematics.

A computation $c \in \Gamma$ can be of two different types. The first type is called a \textit{stencil computation}.

\begin{mydef}
A \textit{stencil computation} is defined as a quintuplet $s(R,w,D,\text{exp},\mathcal{N})$, where $R \subset \Delta, w \in \Delta$, $D$ is one the subsets $D_i$ and $w : D \rightarrow V$.
\end{mydef}
In a stencil computation $s$, $\forall d \in D$, the stencil numerical expression $\text{exp}$ is applied such that $w(d) = \text{exp}(R(d),R(\mathcal{N}(d))$. In this work, a stencil computation $s(R,w,D,\text{exp},\mathcal{N})$ always verifies $R \cap w = \emptyset$, otherwize an implicit numerical scheme has to be solve which is over the scope of this paper.

Figure~\ref{} gives an example of a stencil computation $s(R,w,D,\text{exp},\mathcal{N})$, where $\mathcal{M}(V,E)$ is a two dimensional Cartesian mesh. A single domain $D$ is defined in this example and is composed of cells formed by a cycle of four vertices $v \in V$ and four edges $e \in E$. Furthermore, in this example $R=\{A\}$, $w=B$, and for $(x,y) \in D$ the neighborhood function is 
\begin{equation*}
\mathcal{N} : (x,y) \rightarrow \{(x,y+1),(x,y-1),(x+1,y),(x-1,y)\}.
\end{equation*}
Finally, the numerical expression of this example is 
\begin{equation*}
\text{exp}(A(x,y),A(\mathcal{N}(x,y)) = B(x,y) = A(x,y)+(A(x,y+1)+A(x,y-1)+A(x+1,y)+A(x-1,y))/4.
\end{equation*}

Finally, the second type of numerical computation is a local computation.
\begin{mydef}
A local computation is a quadruplet $l(R,w,D,\text{exp})$, where $e$ does not involve a neighborhood function $\mathcal{N}$.
\end{mydef}

%-------------------------------------
\subsection{Parallelization}
\label{sect:parall}
Multi-stencil mesh-based numerical simulation can be parallelized in various ways and is an interesting kind of application to take advantage of modern heterogeneous HPC architectures, mixing clusters, multi-cores CPUs, vectorization units, GPGPU and many-core accelerators.

\paragraph{Coarse-grain data parallelism.} In a data parallelization technique, the idea is to split the data on which the program is computed in balanced sub-parts, one for each available resource. The same sequential program can afterwards be applied on each sub-part simultaneously, with some additioinal synchronizations between resources to update the data not computed locally and thus to guarantee a correct result.

More formally, the data parallelization of a multi-stencil program $\mathcal{MSP}(T,\mathcal{M},\Delta,\Gamma)$ consists in, first, a partitioning of the mesh $\mathcal{M}$ in $p$ balanced sub-meshes (for $p$ resources) $\{\mathcal{M}_0,\dots,\mathcal{M}_{n-1}\}$. This step can be performed by an external graph partitionner~\cite{} and is not adressed by this paper. As a data is mapped onto the mesh, the set of data $\Delta$ is partitionned the same way than the mesh in $\{\Delta_0,\dots,\Delta_{n-1}\}$. The second step of the parallelization is to identify in $\Gamma$ the needed synchronizations between resources to update data, and thus to build a new ordered list of computations $\Gamma'$.

\begin{mydef}
For $n$ the number of computations in $\Gamma$, and $i<j<n$, a \textit{synchronisation} is needed between $c_i$ and $c_j$, denoted $c_i \ll c_j$, if $c_j=s_j(R_j,w_j,D_j,\text{exp}_j)$, and $w_i \cap R_j \neq \emptyset$. Moreover, the data to update is $w_i \cap R_j$.
\end{mydef}

Actually, a synchronization can only be needed by the data read in a stencil computation, and only if this data has been modified before, which means that it has been written before. This synchronization is needed because the neighborhood function $\mathcal{N}$ of the stencil computation involves values computed on different resources.

\begin{mydef}
A synchronization between two computations $c_i \ll c_j$ is defined as a specific computation $\text{update}(w_i \cap R_j)$.
\end{mydef}

We denote an function to insert at a precise position of a given ordered list, a computation $c_{up}$, as
\begin{equation}
\text{insert} : [c_k]_{0 \leq k \leq n-1} \times \text{position} \times c_{up} \rightarrow [c_0,\dots,c_{\text{position}-1},c_{up},\dots,c_{n-1}]
\end{equation}

The new ordered list of computations $\Gamma'$ is obtained from the insertion of the synchronizations at the good position. Algorithm~\ref{alg:gammaprime} illustrates how $\Gamma'$ is build.
\begin{center}
\begin{algorithm}
$\Gamma' = \Gamma$\\
$\text{offset} = 0$\\
\For{$c_j \in \Gamma$}{
	\uIf{$c_j$ is a stencil}{
	\uIf{$\Gamma[c_0:c_j]==[]$}{
	$\text{insert}(\Gamma',0,\text{update}(R_j)$\\
	$\text{offset}++$\\
	}
	\uElse{
	\For{$c_i \in \Gamma[c_0:c_j]$}{
	\uIf{$c_i \ll c_j$}{
	$\text{insert}(\Gamma',j+\text{offset},\text{update}(w_i \cap R_j)$\\
	$\text{offset}++$
	}
	}
	}
	}
	}
\caption{Creation of $\Gamma'$ from $\Gamma$}
\label{alg:gammaprime}
\end{algorithm}
\end{center}

 The final step of this parallelization is to run $\Gamma'$ on each resource. Thus, for each resource $0 \leq k \leq p-1$ a multi-stencil program defined by
\begin{equation}
\mathcal{MSP}_k(T,\mathcal{M}_k,\Delta_k,\Gamma'),
\end{equation}
is runned.

We denote this parallelization technique a coarse-grain data parallelization in contrast with the same technique applied to the finer level of a single computation. In this case, for a computation $c(R,w,D,\text{exp})$, the local domain $D$ is divided for $p$ resources $\{D_0,\dots,D_{p-1}\}$, and each resource $k$ is responsible for a sub-computation $c_k(R,w,D_k,\text{exp})$.
This finer data parallelization does not require the addition of synchronizations because each computation . This finer data parallelization technique is not directly adressed by the work presented in this paper as it will be explained later.

\paragraph{Task parallelism.}
